---
title: "Web-scraping with rvest"
description: |
  Websraping with R and rvest is very convenient
author:
  - name: Sitendu Goswami
    url: {}
date: 2022-05-14
output:
  distill::distill_article:
    self_contained: false
---

The Internet is filled with a lot of information and as data enthusiasts it is our sacred duty to get data from wherever we can. Some websites provide access to their data through apis, while others are not as forthcoming. First we load the packages

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(rvest)
library(httr)
library(htmltools)
library(stringr)
library(tidyverse)
```

There is a nice website called [91wheels](https://www.91wheels.com), that helps us look at car prices and compare between them. We shall use this website to gather updated car prices, which we shall use for downstream analyses.

The first step is to identify the website and copy its entire url.
Next, we use the "read_html" function to read the entire webpage.
This is how webscraping works, since HTML is a language to display content on webpages, it has tags for all relevant content that we see anywhere. Since most of the times, the data of our interest is in text format, we can gather the relevant information by reading the entire webpage.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tiago<- read_html("https://www.91wheels.com/cars/between-5-lakh-to-10-lakh")
```

In the next step, I use the selector gadget in chrome extension market to find out the xml path to the two variables of my target i.e. car name and car price range.


```{r, message=FALSE, warning=FALSE, echo=TRUE}
price<- tiago |> html_nodes("._cardofcar_price__3qcNH") |> html_text() 
price1<- price |> str_remove_all("[:alpha:]") |> str_remove_all("^\\W") |> as_tibble()
name<- tiago |> html_nodes("._cardofcar_titleName__2ZokJ") |> html_text() |>  as_tibble()
```

Now we have two tibbles called "name" and "price1" that have all the relevant information we need to create the dataset.
But if we look at the two tibbles, we can see that we are long way from clean data that is ready for analysis

```{r}
head(name) # This tibble is fine since car names are character variables
head(price1) # We want a single car price without any other strings and annoyiung characters like the rupee symbol
```



```{r, message=FALSE, warning=FALSE, echo=TRUE}

price1<- price1 |> separate(value, into= c("lower", "higher"), sep = "-")
head(price1)
```

Now we join the car names with the prices to get a data set that is shaping up nicely. We can use the "bind_cols" or "cbind"  functions to join them.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
final<- bind_cols(name, price1)
head(final)
```
Still the pesky rupee symbol continues to pursue us.
Notice there are spaces around the rupee symbol which we can use to separate the column into multiple columns and then select the column which only contains the numeric vector of price.
Moreover, the higher and lower prices of the car should have been changed to a numeric vector by now.


```{r, message=FALSE, warning=FALSE, echo=TRUE}
final_prices<- final |> separate(lower, into = c("symbol", "lower", "space"), sep = " ") |> select(value,lower, higher )|> 
  separate(higher, into = c("symbol", "space", "higher"), sep = " ") |> 
  select(value,lower, higher ) |> 
  mutate(lower = as.double(lower),
         higher = as.double(higher))
head(final_prices)
final_prices |> knitr::kable(type = "html") # our final output looks like this.
```
### Scrape the Decathlon India website

```{r}
decathlon<- function(search, pagenum){
  library(rvest)
  library(tidyverse)
  page<-("https://www.decathlon.in/search?query=")
  pagecont<- paste0(page, search)
  prices<- pagecont |>
    read_html() |>
    html_nodes("._3wHKeni9X-") |>
    html_text()
  # prices
  productname<- pagecont |> read_html() |>
    html_nodes(".card-title") |>
    html_text()
  # productname

  price<- tibble(productname, prices)
  price2<- bind_rows(NULL, price)
  print(price2)
}

```

```{r}
data<- decathlon("merino", 1) |> janitor::clean_names()
head(data)
```


```{r, eval = false}
library(rvest)
library(httr)
link <- read_html("https://www.olx.in/dehradun_g4059236/cars_c84?filter=first_owner_eq_1%2Cmake_eq_volkswagen%2Cpetrol_eq_petrol%2Cyear_between_2015_to_2021")
link |>  html_nodes("._3GOwr") |>  html_text()
```

